The Magnetorotational Instability Prefers Three Dimensions

### 2019-07-23 paper/paper.tex
More words. Acknowledgements. Removed onecolumngrid for fig 1
* this fixes the weirdness after the long equation. Can be put back
  later for more accurate length estimation.

Here, we investigate the linear stability of three dimensional perturbations near the two dimensional critical shear rate $S_c$ for a nearly invisicd, ideal MHD flow.
We find that the first modes to become destabilized are three-dimensional and thus act as a dynamo even in the absence of secondary instability.
These results strongly suggest that the non-normality of the linear operator driving the MRI are always important, even when axisymmetric modes dominate.
% \begin{itemize}
% \item 3D $\to$ dynamo
% \item stars, NSSL
% \item experiments
% \item Squire's theorem violation? Check MHD Squire, check S instead of Re...
% \item Do eigenvalue solution for B growth? i.e.\ check if it is a dynamo?
% \end{itemize}

@@

Figure~\ref{fig:growth_rate}b compares the numerically growth rates for $\SSC=1.002$ between $0 \le k_y \le 0.2$ to this asymptotic approximation, showing that the numerical calculations accurately reproduce the analytic ones in the region where the latter is valid.
We provide the full spectrum for the most unstable parameter values for $\SSC = 1.02$ in figure~\ref{fig:growth_rate}c.
The single unstable mode is purely growing (orange dot), and its complex conjugate is also present (blue dot at $\omega = 0$), consistent with equation~(\ref{eq:asymp}).
The other stable modes (blue dots) are rotationally modified Alfven waves found in left- and right- going pairs, consistent with the analytic predictions for two-dimensional stability calculations.
%
\onecolumngrid

\begin{figure*}[h!]
  \includegraphics[width=\textwidth]{fig_1.pdf}
  \caption{Stuff}
  \label{fig:growth_rate}
\end{figure*}
\twocolumngrid
%
Numerical simulations have consistently shown axisymmetric modes dominating early evolution of the MRI before breaking down into MHD turbulence \citep{1995ApJ...440..742H,2018ApJ...853..174H,2019ApJS..241...26D}. 
These axisymmetric modes, called ``channel modes,'' are exact non-linear solutions of the governing equations with shearing-periodic boundary conditions in $x$, and thus must saturate via parasitic modes from physics beyond non-linear mode coupling \citep{1994ApJ...432..213G}.
Though we use impenetrable, stress-free boundary conditions that do not allow the unbounded growth of channel modes, the basic principle that axisymmetric disturbences grow fastest is not clearly caused by boundary conditions.
thus, we must reconcile our results with those earlier simulations.
In figure~\ref{fig:alpha}, we plot the angle $\alpha = \arctan k_y/k_z$ of the fastest growing mode with the $z$ axis as function of $\SSC$.
Above $\SSC \gtrsim 2.05$, $\alpha$ is zero, indicating that axisymmetric modes have the fastest growth rates.
For the fiducial run in \citet{1996ApJ...464..690H}, $\SSC \simeq 4.84$ (similar values are found in most works).
Thus, our theory predicts that the linear dynamics should be dominated by axisymmetric modes for the parameters studied in prior numerical simulations.
%
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{alpha_vs_ssc_grid.pdf}
  \caption{Angle $\alpha$ vs $\SSC$. Stability limit is at $\SSC \simeq 0.102$}
  \label{fig:alpha}
\end{figure}
%

@@

% If you have acknowledgments, this puts in the proper section head.
%\begin{acknowledgments}
% put your acknowledgments here.
%\end{acknowledgments}

>>>>

Here, we investigate the stability of three dimensional perturbations near the two dimensional critical shear rate $S_c$ for a nearly invisicd, ideal MHD flow.
We find that the first modes to become destabilized are three-dimensional and thus act as a dynamo even in the absence of secondary instability.
These results strongly suggest that the non-normality of the linear operator driving the MRI are always important, even when axisymmetric modes dominate.
% \begin{itemize}
% \item 3D $\to$ dynamo
% \item stars, NSSL
% \item experiments
% \item Squire's theorem violation? Check MHD Squire, check S instead of Re...
% \item Do eigenvalue solution for B growth? i.e.\ check if it is a dynamo?
% \end{itemize}

@@

Figure~\ref{fig:growth_rate}b compares the numerically growth rates for $\SSC=1.002$ between $0 \le k_y \le 0.2$ to this asymptotic approximation, showing that the numerical calculations accurately reproduce the analytic ones in the region where the latter is valid.
We provide the full spectrum for the most unstable parameter values for $\SSC = 1.02$ in figure~\ref{fig:growth_rate}c.
The single unstable mode is purely growing (orange dot), and its complex conjugate is also present (blue dot at $\omega = 0$), consistent with equation~(\ref{eq:asymp}).
The other stable modes (blue dots) are rotationally modified Alfven waves found in left- and right- going pairs, consistent with the analytic predictions for two-dimensional stability calculations.
%
%\onecolumngrid

\begin{figure*}[h!]
  \includegraphics[width=\textwidth]{fig_1.pdf}
  \caption{Growth rates for three-dimensional MRI modes. a) growth rates $\gamma$ over a grid of $k_y$ and $k_z$ for four values of supercriticality $\SSC$. The white contour highlights $\gamma = 0$; the red dots indicate the fastest growing mode. At $\SSC = 0.640$, there are no two dimensional modes. b) Growth rate vs $k_y$ for $k_z = ???$ at $\SSC = 1.002$ (highlighted by the blue line in a). The orange line gives the asymptotic result from equation~(\ref{eq:asymp}) and the blue line are the numerical results. The asymptotic form is valid as $\SSC \to 1$ and $k_y \ll 1$. c) The full discrete spectrum of the MRI for $(k_y, k_z) \simeq (0.263, 0.447)$. The unstable mode is shown in orange; all other modes are stable ($\gamma \le 0$).}
  \label{fig:growth_rate}
\end{figure*}
%\twocolumngrid
%
Numerical simulations have consistently shown axisymmetric modes dominating early evolution of the MRI before breaking down into MHD turbulence \citep{1995ApJ...440..742H,2018ApJ...853..174H,2019ApJS..241...26D}. 
These axisymmetric modes, called ``channel modes,'' are exact non-linear solutions of the governing equations with shearing-periodic boundary conditions in $x$, and thus must saturate via parasitic modes from physics beyond non-linear mode coupling \citep{1994ApJ...432..213G}.
Though we use impenetrable, stress-free boundary conditions that do not allow the unbounded growth of channel modes, the basic principle that axisymmetric disturbences grow fastest is not clearly caused by boundary conditions.
thus, we must reconcile our results with those earlier simulations.
In figure~\ref{fig:alpha}, we plot the angle $\alpha = \arctan k_y/k_z$ of the fastest growing mode with the $z$ axis as function of $\SSC$.
Above $\SSC \gtrsim 2.05$, $\alpha$ is zero, indicating that axisymmetric modes have the fastest growth rates.
For the fiducial run in \citet{1996ApJ...464..690H}, $\SSC \simeq 4.84$ (similar values are found in most works).
Thus, our theory predicts that the linear dynamics should be dominated by axisymmetric modes for the parameters studied in prior numerical simulations.
%
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{alpha_vs_ssc_grid.pdf}
  \caption{Angle $\alpha$ of mode with respect to $z$ axis vs $\SSC$. The inset shows the transition from axisymmetric modes to three dimensional modes as $\SSC$ decreases below $2.05$. The shaded region indicates stability to all perturbations; the stability limit for three dimensional modes is $\SSC \simeq 0.102$.}
  \label{fig:alpha}
\end{figure}
%

@@

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
We would like to thank Steve Tobias for helpful conversations on this work.
JSO acknowledges funding from NASA LWS grant No. NNX16AC92G and Research Corporation Scialog Collaborative Award (TDA) ID\#24231. Computations were performed on the \emph{Leavitt} cluster at the Bates College High Performance Computing Center.
Morgan Baxter acknowledges support from NASA LWS grant No. NNX16AC92G.
\end{acknowledgments}
